---
title: "DSC680 Project2 - Vet Hospital Wait Times - Part1"
author: "Amie Davis"
date: "23 April, 2021"
output:
  pdf_document: default
  word_document: default
---
# Data Sources:
Dove Lewis Animal Hospital, Portland, OR, 1Jan2019-11Apr2021, Proprietary Data

# Load Libraries

```{r libs, message=FALSE}
library(readxl)
library(readr)
library(ggplot2)
library(lubridate)
library(dplyr)
#require(scales)
```

# 1. Prepare Data

## a) Import the Data
```{r load, message=FALSE, warning=FALSE}

# Column range required for all sheets 

# Load Whiteboard data
board_df <- read_excel("Data/WhiteBoard Tracker.xlsx", range = cell_cols("A:K"))

# Load Wait data
# Not loading correctly with read_excel, so switched to read_csv and reformatted dates
wait_df <- read_csv("Data/Smart Flow Wait Times.csv",  
                    col_types = cols_only(
#                                          "Clinical ID" = col_double(),
                                          "Time Stamp" = col_datetime(format="%m/%d/%y %H:%M"),
                                          "Patient Count" = col_integer(),
                                          "Wait Time Average" = col_double(),
                                          "Time Block" = col_integer(),
                                          "Window" = col_character()
                                          ))

# Load Patient data
# Converted Excel file to csv using Trifacta to split data fields that were delimited
patient_df <- read_csv("Data/Patient Snapshot_Revised.csv",
                         col_types = cols_only("Department" = col_character(),
                                          "Consult Date" = col_date(format="%m/%d/%Y"),
                                          "Consult Division" = col_character(),
                                          "Clinical Number" = col_integer(),
                                          "Patient Number" = col_integer(),
                                          "Triage Type" = col_integer(),
                                          "Clinical Description" = col_character(),
                                          "Appointment Type1" = col_character(),
                                          "Appointment Type2" = col_character(),
                                          "Appointment Type3" = col_character(),
                                          "Appointment Date1" = col_datetime(format="%m/%d/%Y %H:%M"),
                                          "Appointment Date2" = col_datetime(format="%m/%d/%Y %H:%M"),
                                          "Appointment Date3" = col_datetime(format="%m/%d/%Y %H:%M"),
                                          "Presenting Problem1" = col_character(),
                                          "Presenting Problem2" = col_character(),
                                          "Presenting Problem3" = col_character(),
                                          "Therapeutic-Procedure1" = col_character(),
                                          "Therapeutic-Procedure2" = col_character(),
                                          "Therapeutic-Procedure3" = col_character(),
                                          "Therapeutic-Procedure4" = col_character(),
                                          "Therapeutic-Procedure5" = col_character(),
                                          "Therapeutic-Procedure6" = col_character(),
                                          "Therapeutic-Procedure7" = col_character(),
                                          "Therapeutic-Procedure8" = col_character()
                                          ))
```

## b) Review Features

```{r bstr, eval=FALSE}
str(board_df)
```

```{r wstr, eval=FALSE}
str(wait_df)
```

```{r pstr, eval=FALSE}
str(patient_df)
```

## c) Derived Features

```{r derived}
# Convert date fields to date stamps
# Split date fields into separate columns using lubridate package
# Time Stamp is POSIXlt, so need to convert to date first
board_df$Date_Stamp <- date(board_df$"Time Stamp")
board_df$TS_HOUR <- hour(board_df$"Time Stamp")
board_df2 <- board_df %>% mutate(Date_Stamp = ymd(Date_Stamp))
board_df_new <- board_df2 %>% mutate (TS_YEAR = year(Date_Stamp),
                                      TS_MONTH = month(Date_Stamp), 
                                      TS_DAY = day(Date_Stamp),
                                      TS_DOW = weekdays(Date_Stamp),
                                      TS_WEEK = week(Date_Stamp)
                                      )

wait_df$Date_Stamp <- date(wait_df$"Time Stamp")
wait_df$TS_HOUR <- hour(wait_df$"Time Stamp")
wait_df2 <- wait_df %>% mutate(Date_Stamp = ymd(Date_Stamp))
wait_df_new <- wait_df2 %>% mutate (TS_YEAR = year(Date_Stamp),
                                      TS_MONTH = month(Date_Stamp), 
                                      TS_DAY = day(Date_Stamp),
                                      TS_DOW = weekdays(Date_Stamp),
                                      TS_WEEK = week(Date_Stamp)
                                      )

# Convert categorical variables to factors
board_df_new$TS_YEAR <- factor(board_df_new$TS_YEAR)
board_df_new$TS_MONTH <- factor(board_df_new$TS_MONTH)
board_df_new$TS_DAY <- factor(board_df_new$TS_DAY)
board_df_new$TS_HOUR <- factor(board_df_new$TS_HOUR)
board_df_new$TS_DOW <- factor(board_df_new$TS_DOW)

wait_df_new$TS_YEAR <- factor(wait_df_new$TS_YEAR)
wait_df_new$TS_MONTH <- factor(wait_df_new$TS_MONTH)
wait_df_new$TS_DAY <- factor(wait_df_new$TS_DAY)
wait_df_new$TS_HOUR <- factor(wait_df_new$TS_HOUR)
wait_df_new$TS_DOW <- factor(wait_df_new$TS_DOW)
wait_df_new$"Time Block" <- factor(wait_df_new$"Time Block")
wait_df_new$Window <- factor(wait_df_new$Window)

patient_df$Department <- factor(patient_df$Department)
patient_df$"Consult Division" <- factor(patient_df$"Consult Division")
patient_df$"Triage Type" <- factor(patient_df$"Triage Type")
patient_df$"Therapeutic-Procedure1" <- factor(patient_df$"Therapeutic-Procedure1")
patient_df$"Therapeutic-Procedure2" <- factor(patient_df$"Therapeutic-Procedure2")
patient_df$"Therapeutic-Procedure3" <- factor(patient_df$"Therapeutic-Procedure3")
patient_df$"Therapeutic-Procedure4" <- factor(patient_df$"Therapeutic-Procedure4")
patient_df$"Therapeutic-Procedure5" <- factor(patient_df$"Therapeutic-Procedure5")
patient_df$"Therapeutic-Procedure6" <- factor(patient_df$"Therapeutic-Procedure6")
patient_df$"Therapeutic-Procedure7" <- factor(patient_df$"Therapeutic-Procedure7")
patient_df$"Therapeutic-Procedure8" <- factor(patient_df$"Therapeutic-Procedure8")
patient_df$"Appointment Type1" <- factor(patient_df$"Appointment Type1")
patient_df$"Appointment Type2" <- factor(patient_df$"Appointment Type2")
patient_df$"Appointment Type3" <- factor(patient_df$"Appointment Type3")
patient_df$"Presenting Problem1" <- factor(patient_df$"Presenting Problem1")
patient_df$"Presenting Problem2" <- factor(patient_df$"Presenting Problem2")
patient_df$"Presenting Problem3" <- factor(patient_df$"Presenting Problem3")


head(board_df_new)
head(wait_df_new)
head(patient_df)
```

## d) Summary Statistics
```{r bsumm, eval=FALSE}
#str(board_df_new)
summary(board_df_new)
```
Observations: \newline\
181,742 Whiteboard Records \newline\
No N/A values or outliers \newline\

```{r wsumm, eval=FALSE}
#str(wait_df_new)
summary(wait_df_new)
```
Observations: \newline\
64,032 Smart Flow Wait Records \newline\
No N/A values \newline\
Extreme estimated wait times confirmed by source \newline\

```{r psumm, eval=FALSE}
str(patient_df)
#summary(patient_df)
```
Observations: \newline\
74,529 Weekly Client Records \newline\
68,291 Records do not have a Triage Type. \newline\
- Hospital did not start tracking Triages until 4 Jan 2021 \newline\
- Triages are tracked for emergency room only. \newline\
26 Records do not have an Appointment Type. \newline\
89 Records do not have an Appointment Date. \newline\
74,422 Records do not have a Presenting Problem recorded. \newline\
56,511 Records do not have a Procedure listed.

## e) Drop unneeded columns

```{r drop, include=TRUE}

# Drop unused pre-calculated datetime fields

board_df_new[ ,c(
  "Time",
  "TIME Hour",
  "Weekday",
  "Date",
  "Year",
  "Week",
  "Month"
)] <- list(NULL)
```

## f) Remove Outliers
```{r outliers, include=TRUE}

# If no 1st appt date, use 2nd appt date
# Note that ifelse changes type, so using dplr's if_else
patient_df$Appt_Date <- if_else(is.na(patient_df$"Appointment Date1"), patient_df$"Appointment Date2", patient_df$"Appointment Date1")

# Only use patient records w appt date
clean_data<-subset(patient_df,!(is.na(patient_df["Appt_Date"])))

# Only use patient records w appt type
clean_data<-subset(clean_data,!(is.na(clean_data["Appointment Type1"])))

# Look specific record number 901362
#test_data<-subset(clean_data,clean_data$"Clinical Number" == 901362)
#head(test_data)

#summary(clean_data)
#str(clean_data)
#74,479 records remaining
```

## g) Export dataset
```{r exp, include=TRUE}

# Export data for use in visualizations
write.csv(clean_data,"Data/patient_new.csv")

```

# 2. EDA - Review Distributions

## a) Plot Histograms for Numeric Vars

```{r date_hist}

ggplot(clean_data, aes(x=Appt_Date)) +
    geom_histogram(bins=150) +
    labs(x="Appointment Date") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
\newline\
There appears to be a peak season every year.

## b) Histograms for Categorical Features
```{r dep_plot}

# Department
p <- ggplot(clean_data, aes(x=Department)) +
     geom_bar(fill="dark blue") +
     labs(x="Department", y="Records", title="Records By Department") +
     theme(axis.text.x = element_text(angle = 90, hjust = 1))

p
```
\newline\
As you can see, most appointments are in the Emergency Critical Care Department.  This includes both Outpatient and ICU patients.

```{r div_plot}

# Offense Codes

p <- ggplot(clean_data, aes(x=clean_data$"Consult Division")) +
     geom_bar(fill="dark blue") +
     labs(x="Consult Division", y="Records", title="Records By Consult Division") +
     theme(axis.text.x = element_text(angle = 90, hjust = 1))

p
```
\newline\
This is expected.  Most appointments are in the Emergency Critical Care Division.  
